Efficient forwards-time simulation with tree sequences

### 2017-11-12 forwards_paper.tex
.
more edits

%% Merged this above; moved the "hybrid approach" to the conclusion.
% Another critical assumption of the coalescent is that the sample size $n$ is much smaller
% than the effective population size $N_e$. 
% With the very large sample sizes becoming common
% in human genetics, this is no longer a safe assumption. 
% For example, a recent study~\citep{martin2017human}
% simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
% reference panels in genome wide association studies. While many results of the
% coalescent are surprisingly
% robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
% are distorted and coalescent simulations may be
% misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
% A hybrid approach, in which the recent history of a large sample of
% individuals is simulated under a detailed forwards-in-time model and the deep history
% under the coalescent, is an attractive prospect.
% \jk{Not sure where this should go; perhaps in the discussion? We can make the point that
% by keeping track of the genealogies forwards in time we can easily combine these with the
% deep history simulated by msprime}

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 wieth ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site, 
and results in a specific derived state. 
Thus, all samples below the mutation node in the tree will inherit this state 
(unless further mutations are encountered). 
Three mutations are shown in Figure~\ref{fig:example_tree_sequence}, 
illustrated by red hexagons. 
The first mutation occurs at site zero (in the left-hand tree), and is a simple
mutation resulting in node $2$ inheriting the state 'T'. 
The second site (in the right hand tree) has two mutations: 
one occurring over node $3$ changing the state to `C', 
and a back mutation over node $1$ changing the state to `G'.

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million trees
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB --
a compression factor of 200,000.
Working with either the Newick or VCF encoding
of this dataset would likely require several
days of CPU time simply to read the information info memory.

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
with overlapping generations that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we assume there is exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.

Let $\randomuniform(A)$ to be an element of the set $A$ chosen
uniformly at random (note that each instance of $\randomuniform(A)$ within
an algorithm listing represents an independent draw from the set $A$).

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Runs a forwards time Wright--Fisher simulation of a populations with $N$
individuals with a probability of death per generation $\delta$ and a chromosome of length
$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
describing the recorded edges.
}

@@

\algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
Set $P'_j \leftarrow P_j$ for $0\leq j < N$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
}

@@

\jk{I haven't checked this carefully. If we used 1-based indexes it might make things a bit
simpler}.
\end{taocpalg}

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
For example, if we assume that infinite sites mutations
occur at rate $\mu$ per generation then there are $\mbox{Poisson}\left((r - \ell)(\tau_p
- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
It is straightforward to record this information during the simulation, but it
is significantly simpler and more efficient to generate these mutations
after the simulation has completed. In the case of infinite sites mutations
there is no difference between generating these mutations during the forwards
in time simulation and after it has completed. We are using precisely the same
information.

@@

It is significantly more efficient to generate mutations as a separate process
after the topological simulation has completed because we will generate many
mutations that will be lost in the population. Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright-Fisher
process like Algorithm~\algref{W}. There are two different trees in this
figure: one shows all the edges output by the simulation and the other
is the minimal tree representing the ancestry of the currently alive
population. Clearly there is a great deal of redundancy in the topological
information output by the simulation.

There are two sources of redundancy here. The first type of redundancy arises
from nodes in tree that have only one child. In Algorithm~\algref{W} we do
not attempt to track coalescence events but simply record all parent-child
relationships in the history of the population. As such, my of these edges
will record the simple passing of genealogical information from parent to child
and only some small subset will correspond to coalesences within the marginal
trees. The second source of redundancy in the output of Algorithm~\algref{W}
is due to the fact that lineages die out: there will be a large number of
individuals in the simulation that leave no ancestors in the present day
population. Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
ancestors in the current population and so the entire path tracing back to
the root is redundant.

One of the reasons that it is more efficient to generate mutations on the
genealogies after we have completed the forwards in time simulation is
because we avoid the cost of generating mutations on these dead-end branches.
It is only with the benefit of hindsight that we can know which lineages end
up being ancestral to the sample, and we can avoid a substantial cost of
generating and maintaining mutational information on these evolutionary
dead-ends. Assuming mutations are rare, another  reason that it is more efficient
to place mutations on the trees afterwards is that we have far fewer edges
in the trees: because we have removed all the intermediate `unary' nodes, we
have fewer edges to generate mutations on.

There are many advantages to having minimal genealogies such as show in
Figure~\ref{fig:wf-trees}. Computing this minimal representation of the
edges output by a forward-time simulation is an instance of a more
general problem that we refer to as `tree sequence simplification'. We discuss
this problem and an efficient solution in the next subsection.

>>>>

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 with ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site, 
and results in a specific derived state. 
Thus, all samples below the mutation node in the tree will inherit this state 
(unless further mutations are encountered). 
Three mutations are shown in Figure~\ref{fig:example_tree_sequence}, 
illustrated by red hexagons. 
The first mutation occurs at site zero (in the left-hand tree), and is a simple
mutation resulting in node $2$ inheriting the state 'T'. 
The second site (in the right hand tree) has two mutations: 
one occurring over node $3$ changing the state to `C', 
and a back mutation over node $1$ changing the state to `G'.

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million trees
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB --
a compression factor of 144,000.
Working with either the Newick or VCF encoding
of this dataset would likely require several
days of CPU time simply to read the information info memory.

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
with overlapping generations that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random 
(and each such instance represents and independent draw).

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals,
with a probability of death per generation $\delta$ and a chromosome of length
$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
describing the recorded edges.
}

@@

\algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
Set $P' \leftarrow P$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
}

@@

\end{taocpalg}

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
For example, if we assume that mutations occur according to the infinite sites model
at rate $\mu$ per generation per unit length then there are 
$\mbox{Poisson}\left(\mu (r - \ell)(\tau_p
- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be associated with node $c$.
It is straightforward to record this information during the simulation, but it
is significantly simpler and more efficient to generate these mutations
after the simulation has completed. 
% In the case of infinite sites mutations
% there is no difference between generating these mutations during the forwards
% in time simulation and after it has completed. We are using precisely the same
% information.

@@

Besides the cost of transmitting genotypes,
generating mutations as a separate process after the topological simulation has completed 
avoids generating the many mutations that are lost in the population. 
Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright--Fisher
process like Algorithm~\algref{W}. 
On the left is the tree showing all the edges output by the simulation, 
while on the right
is the minimal tree representing the ancestry of the currently alive
population. Clearly there is a great deal of redundancy in the topological
information output by the simulation.

There are two sources of redundancy here. The first type of redundancy arises
from nodes in tree that have only one child. In Algorithm~\algref{W} we do
not attempt to track coalescence events but simply record all parent-child
relationships in the history of the population. As such, many of these edges
will record the simple passing of genealogical information from parent to child
and only some small subset will correspond to coalesences within the marginal
trees. The second source of redundancy in the output of Algorithm~\algref{W}
is due to the fact that lineages die out: a large number of
individuals in the simulation leave no descendants in the present day population. 
Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
ancestors in the current population, and so the entire path tracing back to
the root is redundant.

A minimal genealogy such as in Figure~\ref{fig:wf-trees}
has fewer edges on which to place mutations.
There are many more advantages to simplifying the genealogies to this point.
Computing this minimal representation of the
edges output by a forward-time simulation is an instance of a more
general problem that cover in the next section.
